Detection of Treatment-Subgroup Interactions in Clustered Datasets with Generalized Linear Mixed-effects Model (GLMM) Trees

### 2015-08-18 pkg/glmertree/paper/treatment_interactions paper.tex
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@747 255f4249-2263-4417-b1ea-8498779e27a7

\documentclass[nobf,doc]{apa}
\usepackage{amssymb}
\usepackage{amsmath,amsthm,tabularx,curves,texdraw,psfrag}
\usepackage{apacite}
\usepackage[english]{babel}
\usepackage{lscape,dcolumn,hhline}
\usepackage{graphicx,epic,eepic,rotating}
\usepackage[normalem]{ulem}
\usepackage{fancyhdr}
\usepackage{eurosym,bbding}
\usepackage{verbatim}
\usepackage{ctable}
\usepackage{multirow}
\usepackage{rotating}
\usepackage{lineno}
\usepackage{tikz}
\linespread{1.2}

\usetikzlibrary{shapes,shadows,arrows,positioning}
\title{Detection of Treatment-Subgroup Interactions in Clustered Datasets: Combining Model-Based Recursive Partitioning and Random-Effects Estimation}
\shorttitle{Detection of treatment-subgroup interactions in clustered datasets}
\author{M. Fokkema$^1$, N. Smits$^2$, A. Zeileis$^3$, T. Hothorn$^4$, H. Kelderman$^5$}
\affiliation{$^1$Universiteit Leiden, $^2$Universiteit van Amsterdam, $^3$Universit\"{a}t Innsbruck, $^4$Universit\"{a}t Z\"{u}rich, $^5$Universiteit Leiden and Vrije Universiteit, Amsterdam}

@@

\section{Abstract}
Identification of subgroups of patients for which treatment A is more effective than treatment B, and vice versa, is of key importance to the development of personalized medicine. Several tree-based algorithms have been developed for the detection of such treatment-subgroup interactions. In many instances, however, datasets may have a clustered structure, where observations are clustered within, for example, research centers, studies or persons. In the current paper we propose a new algorithm, GLMMtree, that allows for detection of treatment-subgroup interactions, as well as estimation of cluster-specific random effects. The algorithm uses model-based recursive partitioning (MOB) to detect treatment-subgroup interactions, and a linear mixed-effects model for estimation of random-effects parameters. In a simulation study, we evaluate the performance of GLMMtree and compare it with that of MOB trees without random-effects estimation. In datasets without treatment-subgroup interactions, GLMMtree was found to have a much lower Type I error rate than MOB trees without random effects (4 and 33\%, respectively). Furthermore, in datasets with treatment-subgroup interactions, GLMMtree recovered the true treatment subgroups much more often than MOB without random effects (90\% and 61\% of the datasets, respectively). Also, GLMMtree predicted treatment outcome differences more accurate than MOB trees without random effects (average accuracy of .94 and .88, respectively). We illustrate the application of GLMMtree on a patient-level dataset of a meta-analysis on the effects of psycho- and pharmacotherapy for depression. We conclude that GLMMtree is a promising algorithm for the detection of treatment-subgroup interactions in clustered datasets, and discuss directions for future research.

@@

In medicine-efficacy research, the one-size-fits-all paradigm is slowly losing ground, and personalized medicine is becoming increasingly important. Personalized medicine presents the challenge of finding which patients respond best to which treatments. This can be referred to as the detection of treatment-subgroup interactions \cite<e.g., >{DoovyDuss14}. In most cases, treatment-subgroup interactions are studied using linear models, such as factorial analysis of variance techniques, in which potential moderators have to be specified a-priori, have to be checked one at a time, and continuous moderator variables have to be discretized a-priori. This may hamper identification of which treatments work best for whom, especially when there are no a-priori hypotheses about treatment-subgroup interactions. As noted by \citeA{KreayFran06}, there is a need for methods that generate, instead of test, hypotheses and that are specifically directed at the detection of treatment interactions.

Tree-based methods are such hypothesis-generating methods, as they can automatically detect subgroups which differ on the expected outcomes for one or more treatments. Due to their flexibility, tree-based methods are preeminently suited to the detection of treatment-subgroup interactions: they can handle many potential predictor variables at once, and can automatically detect (higher order) interactions between predictor variables. Several promising tree-based algorithms and software packages have been developed to assist in the detection of treatment-subgroup interactions (e.g., \citeNP{DussyMech14, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08}; see \citeNP{DoovyDuss14} for an overview). Of these tree-based methods, model-based recursive partitioning (MOB; \citeNP{ZeilyHoth08}) may be the most flexible in detecting treatment-subgroup interactions, as it offers a very generic data-analytic framework for detecting partitions in a dataset, with different model parameter estimates. The recursive partitioning in MOB can be based on a broad class of parametric models that can be fitted using M-type estimators \cite{ZeilyHoth08}, the most well-known example being the generalized linear model (GLM). Earlier, MOB has been successfully applied by \citeA{DrieySmit14} in the detection of subgroups with differential treatment outcomes for two different psychotherapies.

However, none of the aforementioned tree-based algorithms allow for taking into account the clustered structure of datasets. In many cases, researchers may want to detect treatment-subgroup interactions in datasets with a clustered structure \cite<e.g., >{KoopyHeij07}. For example, in individual-level patient data meta-analyses, where datasets of multiple trials evaluating the effects of the same treatments are pooled. In such analyses, the clustered structure of the dataset should be taken into account by including study-specific effects in the model, prompting the need for modeling random effects \cite<e.g., >{Frie93,DersyLair86,HiggyWhit01}. Likewise, longitudinal datasets, and datasets from multi-center trials also require modeling of random effects. Ignoring the clustered structure of datasets may lead to biased inference, due to underestimated standard errors \cite<e.g., >{BrykyRaud92,Hox98,NooryOpde05}. When the interest is in subgroup detection, ignoring random effects on the outcome variable may result in the detection of spurious subgroups \cite<e.g., >{SelaySimo12}.

In the current paper, we present a tree-based algorithm for treatment-subgroup interaction detection, which takes the clustered nature of datasets into account. The algorithm combines MOB with the estimation of random effects, thus allowing for the detection of treatment-subgroup interactions, as well as accounting for variation between clusters (e.g., trials). In what follows, we first discuss the existing frameworks for estimating treatment effects: the GLM, model-based recursive partitioning of the GLM, and the generalized linear mixed-effects model (GLMM). Then, we present a new algorithm, which combines model-based recursive partitioning and random-effects estimation: GLMMtree. In a simulation study, we evaluate the comparative accuracy of the new algorithm. Finally, we apply GLMMtree to an existing dataset on the effects of treatments for depression, to illustrate the application of the algorithm.

@@

\subsection{GLM}
In a clinical trial, where the outcomes of two or more treatments are compared, an overall GLM may be used to estimate treatment effects. The goal is to estimate a model for predicting the value of treatment outcome $y$ for observation $i=1,\dots,N$ (an overview of notation used is provided in the Appendix):

\begin{equation}
\label{eq:fixedeffects}
	g(\mu_{i}) = g(E(y_{i})) = x_{i}\beta
\end{equation}

Where $y_i$ is the value of the linear predictor of the response variable for observation $i$, and $g$ is the link function, characterizing the relationship between the linear predictor and the mean of the response distribution function. In case of a continuous response variable, $g$ is the identity function. Further, $x_{i}$ is a vector of fixed-effects predictor variable values for observation $i$, of which the first element takes a value of 1 for the intercept, and the second element takes the value of a dummy indicator for treatment type. $\beta$ is a vector of fixed-effects regression coefficients, containing the intercept, which is the mean value of the linear predictor in the first treatment group, and the slope, which is the difference in mean values of the linear predictor between the first and second treatment groups.

To keep notation and examples simple, we assume $x_{i}$ and $\beta$ to have length 2, That is, the effects of only two treatment conditions are estimated and no additional covariates are included in the linear model. However, additional treatment conditions and covariates can easily be included.

An example of such a GLM for a continuous outcome variable is graphically represented in Figure \ref{fig:fixedeffects}. The example is based on simulated data. The boxplots in Figure \ref{fig:fixedeffects} show the distribution of the outcome variable (posttreatment depression score) among 150 participants, who were randomly assigned to treatments 1 and 2. Little overall difference between the outcomes of both treatments is suggested, as the slope of the regression line is nearly zero. We shall see that this does not necessarily mean that posttreatment depression score and treatment type are unrelated, as the effect of treatment may be moderated by other variables. Conditional on other variables (i.e., subgroup indicators), the relationship between $x$ and $y$ may vary in strength and/or direction (c.f., Simpson's paradox; \citeNP{Simp51}).

\begin{figure}[!htbp]
    \includegraphics[width=8cm]{Interaction_ex;treedepth1.pdf}
	\caption{Example of a linear fixed-effects model for treatment outcomes (N=150). The dot for treatment 1 represents the first, and the slope of the regression line represents the second element of $\beta$.}
	\label{fig:fixedeffects}
\end{figure}

@@

The rationale behind MOB is that a global model for all observations, like that in Equation \ref{eq:fixedeffects}, may not describe the data well, and when additional covariates are available it may be possible to partition the dataset with respect to these covariates, and find a better model in each cell of the partition \cite{ZeilyHoth08}. This is reminiscent of the classification and regression tree (CART) algorithm of \citeA{BreiyFrie84}, which splits the dataset into subsets, for which the distributions of the outcome variable are most different. Whereas CART trees have constant fits in the terminal nodes, MOB trees have parametric models with one or more predictor variables in their terminal nodes.

To find partitions and better-fitting local linear models, the MOB algorithm tests for parameter instability. When the partitioning is based on a GLM, instabilities are differences in $\hat{\beta}$ across partitions of the dataset, which are defined by one or more additional covariates. To find partitions, the MOB algorithm cycles iteratively through the following steps \cite{ZeilyHoth08}: (1) fit the parametric model to the dataset, (2) test for parameter instability over a set of partitioning variables, (3) if there is some overall parameter instability, split the dataset with respect to the variable associated with the highest instability, (4) repeat the procedure in each of the resulting subsamples.

More specifically, in step (2), to test for parameter instability, the so-called \textit{scores} are computed, using the score function. The expected value of the scores over all observations in a dataset is zero, by definition. Under the null hypothesis of parameter stability, the scores do not systematically deviate from the expected value of zero, when the observations are ordered by the values of a potential partitioning variable $U_k$. To statistically test whether there are systematic deviations of the scores from zero with respect to variable $U_k$, the class of generalized M-fluctuation tests is used \cite{Zeil05,ZeilyHorn07}.

If the null hypothesis of parameter stability in step (2) can be rejected, that is, if at least one of the partitioning variables $U_{k}$ has a p-value for the M-fluctuation test below the pre-specified significance level $\alpha$, the dataset is partitioned into two subsets in step (3). In step (3), a binary partition is created using $U_{k*}$, the variable with the minimal p-value in step (2). The split point for $U_{k*}$ is selected, by taking the value that minimizes the sum of the residual sum of squares in both partitions \cite{ZeilyHoth08}. In step (4), steps (1) through (3) are repeated in each partition, until the null hypothesis of parameter stability can no longer be rejected.

Due to the binary recursive nature of MOB, the resulting partitions can be represented as a tree. If the partitioning is based on the GLM, the result is a GLMtree, which has a local fixed-effects regression model in every $j$th (where $j = 1,...,J$) terminal node of the tree. As a result, in the GLMtree model, the value for $\beta$ depends on terminal node $j$ in which observation $i$ 'falls':

\begin{equation}
\label{eq:fixedeffects_MOB}
	g(\mu_{i}) = g(E(y_{i})) = x_{i}\beta_{j}
\end{equation}

Figure \ref{fig:example_mobtree} provides an example of the GLMtree model in Equation \ref{eq:fixedeffects_MOB}, based on the same data as was used for Figure \ref{fig:fixedeffects}. By using four additional covariates (anxiety questionnaire score, duration of depressive symptoms at baseline, age), MOB partitioned the observations into four subgroups, each with a different estimate for $\beta_j$. The leftmost subgroup in Figure \ref{fig:example_mobtree} represents observations with low duration and low anxiety, for which treatment 1 is much more beneficial then treatment 2 (i.e., lower posttreatment depression scores). The rightmost subgroup represents observations with longer duration of depressive symptoms, for whom treatment 2 is much more beneficial than treatment 1. The two terminal nodes in the middle indicate that for patients with low duration, treatment 1 is only beneficial to those with moderate levels of anxiety (i.e., a values $>10$ and $\leq 12$). Age did not have an effect on treatment outcome, and therefore does not appear as a splitting variable in the tree.

\begin{figure}[!h]
    \includegraphics[width=15cm]{Interaction_ex;treedepth3.pdf}
    \caption{Example of tree representation of model-based recursive partitions, based on the same data as Figure \ref{fig:fixedeffects}. Three additional covariates (anxiety questionnaire score, duration of depressive symptoms at baseline in months and age) were used as potential splitting variables.}
    \label{fig:example_mobtree}
\end{figure}

\subsection{GLMM}
When a dataset contains observations from multiple clusters (i.e., trials, research centers, or individuals in longitudinal datasets), the GLM in Equation \ref{eq:fixedeffects} can be extended to include cluster-specific, or random effects, and the model becomes a GLMM:

\begin{equation}
\label{eq:mixedeffects}
	g(\mu_{i}) = g(E(y_{i})) = x_{i}\beta + z_{i}b_{m}
\end{equation}

Where $z_{i}$ is a vector of random-effects predictor variables values for observation $i$, and $b_{m}$ is the vector of random-effects regression coefficients in cluster $m$, of which observation $i$ is part. Within the linear mixed-effects model, it is assumed that values of $b_{m}$ are normally distributed, with mean zero and variance $\sigma^{2}_{b_m}$. For simplicity, we assume $z_{i}$ and $b_{m}$ to have length 1 in the current paper; that is, only cluster-specific intercepts are included in the models. However, random-effects covariates can easily be included. The parameters of the GLMM can be estimated with, for example, maximum likelihood (ML) and restricted ML (REML), as described in for example \cite{BrykyRaud92}. Note that for prediction of the treatment outcome for a new observation from a new cluster, only the fixed-effects part of Equation \ref{eq:mixedeffects} would be used.

\subsection{Combining model-based recursive partitioning and random-effects estimation}

As noted earlier, ordinary GLM(M)s are not well suited for the detection of treatment-subgroup interactions, whereas the MOB algorithm is, but does not allow for estimation of random effects. Therefore, we propose the GLMMtree, which combines the GLMM in Equation \ref{eq:mixedeffects} with the GLMtree in Equation \ref{eq:fixedeffects_MOB}:

\begin{equation}
\label{eq:glimmertree}
	g(\mu_{i}) = g(E(y_{i})) = x_{i}\beta_{j} + z_{i}b_{m}
\end{equation}

To estimate the partitions, and values of $\beta_j$ and $b_m$ for this model, we take an iterative approach, alternating between (1) assuming the random-effects coefficients $b_m$ known, allowing for partitioning the datasets and estimating the corresponding $\beta_j$ values; and (2) assuming the partition and corresponding $\beta_j$ values known, allowing for estimation of the random-effects coefficients $b_m$.

In Figure \ref{fig:algorithm}, a schematic representation of the GLMMtree algorithm is presented. The algorithm initializes by setting all values $b_m$ to $0$, since the random-effects (and also the fixed-effects) parts are initially unknown. In every iteration, the GLMtree (i.e., the partition and corresponding fixed-effects coefficients $\beta_j$) and random-effects coefficients $b_m$ are re-estimated. The GLMtree is estimated, given the estimated $b_m$ values from the last iteration, and the $b_m$ values are estimated, given the estimated GLMtree from the current iteration. Iterations are continued until convergence, which is monitored by computing the log-likelihood criterion of the mixed-effects model in Equation \ref{eq:mixedeffects}. A similar approach has been taken by \citeA{HajjyBell11} and \citeA{SelaySimo12}, who added random-effects estimation to CART trees with constant fits, instead of linear models, in the terminal nodes.

\begin{figure}[!htbp]
	\begin{tabular}{ll}
		\hline 
		\multicolumn{2}{l}{\textbf{Algorithm} GLMMtree}\\
		\hline
		&Step 0: Initialize by setting all values $b_m$ to $0$.\\

		&Step 1: Given the current $b_m$ values, partition dataset and estimate corresponding values of $\beta_j$.\\

		&Step 2: Given the current partition and corresponding $\beta_j$ values, estimate $b_m$ values. \\

		&Step 3: Repeat steps 1 and 2 until convergence.\\
		\hline
	\end{tabular}
	\caption{Description of the GLMMtree algorithm}
	\label{fig:algorithm}
\end{figure}

>>>>

\documentclass[nobf,doc]{apa}
\usepackage{amssymb}
\usepackage{amsmath,amsthm,tabularx,curves,texdraw,psfrag}
\usepackage{apacite}
\usepackage[english]{babel}
\usepackage{lscape,dcolumn,hhline}
\usepackage{graphicx,epic,eepic,rotating}
\usepackage[normalem]{ulem}
\usepackage{fancyhdr}
\usepackage{eurosym,bbding}
\usepackage{verbatim}
\usepackage{ctable}
\usepackage{multirow}
\usepackage{rotating}
\usepackage{lineno}
\usepackage{tikz}
\linespread{1.2}
\usepackage{tabto}

\usetikzlibrary{shapes,shadows,arrows,positioning}
\title{Detection of Treatment-Subgroup Interactions in Clustered Datasets with Generalized Linear Mixed-effects Model (GLMM) Trees}
\shorttitle{GLMM Trees}
\leftheader{GLMM Trees}
\rightheader{GLMM Trees}
\author{M. Fokkema$^1$, N. Smits$^2$, A. Zeileis$^3$, T. Hothorn$^4$, H. Kelderman$^5$}
\affiliation{$^1$Universiteit Leiden, $^2$Universiteit van Amsterdam, $^3$Universit\"{a}t Innsbruck, $^4$Universit\"{a}t Z\"{u}rich, $^5$Universiteit Leiden and Vrije Universiteit, Amsterdam}

@@

\section{Abstract}
Identification of subgroups of patients for which treatment A is more effective than treatment B, and vice versa, is of key importance to the development of personalized medicine. Several tree-based algorithms have been developed for the detection of such treatment-subgroup interactions. In many instances, however, datasets may have a clustered structure, where observations are clustered within, for example, research centers, studies or persons. In the current paper we propose a new algorithm, GLMM tree, that allows for detection of treatment-subgroup interactions, as well as estimation of cluster-specific random effects. The algorithm uses model-based recursive partitioning (MOB) to detect treatment-subgroup interactions, and a generalized linear mixed-effects model for estimation of random-effects parameters. In a simulation study, we evaluate the performance of GLMM tree and compare it with that of MOB trees without random-effects estimation. In datasets without treatment-subgroup interactions, GLMM tree was found to have a much lower Type I error rate than MOB trees without random effects (4 and 33\%, respectively). Furthermore, in datasets with treatment-subgroup interactions, GLMM tree recovered the true treatment subgroups much more often than MOB without random effects (90\% and 61\% of the datasets, respectively). Also, GLMM tree predicted treatment outcome differences more accurately than MOB trees without random effects (average accuracy of .94 and .88, respectively). We illustrate the application of GLMM tree on a patient-level dataset of a meta-analysis on the effects of psycho- and pharmacotherapy for depression. We conclude that GLMM tree is a promising algorithm for the detection of treatment-subgroup interactions in clustered datasets, and discuss some directions for future research.

\textit{Keywords}: model-based recursive partitioning, treatment-subgroup interactions, random effects, generalized linear mixed-effects model, classification and regression trees

%\section{Citation stuff for journal publication}
%
%Potential Psychological Methods references: only \citeA{StroyMall09} (not added yet). Maybe force in: 
%
%''Do logistic regression and signal detection identify different subgroups at risk? Implications for the design of tailored interventions. Kiernan, Michaela; Kraemer, Helena C.; Winkleby, Marilyn A.; King, Abby C.; Taylor, C. Barr. Psychological Methods, Vol 6(1). ''
%
%''Hedges, L. V., \& Pigott, T. D. (2004). The power of statistical tests for moderators in meta-analysis. Psychological methods, 9(4), 426.''
%
%''Cooper, H., \& Patall, E. A. (2009). The relative benefits of meta-analysis conducted with individual participant data versus aggregated data. Psychological methods, 14(2), 165.'' (they write in abstract ''Regardless of whether a meta-analysis is conducted with IPD or AD, synthesists must remain vigilant in how they interpret their results. They must avoid ecological fallacies, Simpson’s paradox, and interpretation of synthesis-generated evidence as supporting causal inferences.'')
%
%''Hedges, L. V., \& Vevea, J. L. (1998). Fixed-and random-effects models in meta-analysis. Psychological methods, 3(4), 486.''
%
%Statistics in Medicine references: \cite{DussyMech14,FostyTayl11,HiggyWhit01,LipkyDmit11}

@@

In research assessing the efficacy of treatments for somatic and psychological disorders, the one-size-fits-all paradigm is slowly losing ground, and personalized medicine is becoming increasingly important. Personalized medicine presents the challenge of finding which patients respond best to which treatments. This can be referred to as the detection of treatment-subgroup interactions \cite<e.g., >{DoovyDuss14}. In most cases, treatment-subgroup interactions are studied using linear models, such as factorial analysis of variance techniques, in which potential moderators have to be specified a-priori, have to be checked one at a time, and continuous moderator variables have to be discretized a-priori. This may hamper identification of which treatments work best for whom, especially when there are no a-priori hypotheses about treatment-subgroup interactions. As noted by \citeA{KreayFran06}, there is a need for methods that generate, instead of test, hypotheses and that are specifically directed at the detection of treatment interactions.

Tree-based methods are such hypothesis-generating methods, as they can automatically detect subgroups which differ on the expected outcomes for one or more treatments. Due to their flexibility, tree-based methods are preeminently suited to the detection of treatment-subgroup interactions: they can handle many potential predictor variables at once and can automatically detect (higher order) interactions between predictor variables. Several promising tree-based algorithms for the detection of treatment-subgroup interactions have been developed (e.g., \citeNP{DussyMech14, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08}; see \citeNP{DoovyDuss14} for an overview). Among these methods, model-based recursive partitioning (MOB; \citeNP{ZeilyHoth08}) seems to be the most flexible tool for detecting treatment-subgroup interactions, as it offers a very generic data-analytic framework for detecting partitions in a dataset, with different model parameter estimates. The recursive partitioning in MOB can be based on a broad class of parametric models that can be fitted using M-type estimators \cite{ZeilyHoth08}, the most well-known example being the generalized linear model (GLM). Earlier, GLM-based MOB has been successfully applied by \citeA{DrieySmit14} in the detection of subgroups with differential treatment outcomes for two different psychotherapies.

However, none of the aforementioned tree-based algorithms allow for taking into account the clustered structure of datasets. In many cases, researchers may want to detect treatment-subgroup interactions in datasets with a clustered structure \cite<e.g., >{KoopyHeij07}. For example, in individual-level patient data meta-analyses, in which datasets of multiple trials evaluating the effects of the same treatments are pooled. In such analyses, the clustered structure of the dataset should be taken into account by including study-specific effects in the model, prompting the need for modeling random effects \cite<e.g., >{Frie93,DersyLair86,HiggyWhit01}. Likewise, longitudinal datasets, and datasets from multi-center trials also require modeling of random effects. Ignoring the clustered structure of datasets may lead to biased inference, due to underestimated standard errors \cite<e.g., >{BrykyRaud92,Hox98,NooryOpde05}. More specifically, when the interest is in subgroup detection, ignoring random effects on the outcome variable may result in the detection of spurious subgroups \cite<e.g., >{SelaySimo12}.

In the current paper, we present a tree-based algorithm for detecting treatment-subgroup interactions, which takes the clustered nature of datasets into account. The algorithm combines MOB with random-effects estimates, thus allowing for the detection of treatment-subgroup interactions, as well as accounting for variation between clusters (e.g., trials). In what follows, we will introduce the existing frameworks for estimating treatment effects: the GLM, model-based recursive partitioning, and the generalized linear mixed-effects model (GLMM). Then, we intoduce a new algorithm, which combines MOB and the GLMM: GLMM tree.

Before we discuss these methods for estimating treatment effects, we will introduce an artificial motivating data set, with which the methods will be illustrated. After we introduce the GLMM tree algorithm, we present a simulation study, in which we evaluate GLMM trees comparative accuracy. Finally, in the application, we use GLMM tree to detect treatment-subgroup interactions in an existing dataset on the effects of treatments for depression.

\subsection{Artificial motivating dataset}

To illustrate the application of the methods to be discussed, we will use a simulated dataset of 150 observations, which were randomly assigned to Treatment 1 (78 observations) or Treatment 2 (72 observations). Every observation has a value on the response variable, with which the effect of treatment is assessed: the posttreatment total score on a depression inventory. Further, all observations have values on three covariates: duration of depressive symptoms prior to treatment in months (range 0-15); age in years (range 18-75); anxiety inventory total score (range 3-18).

The simulated dataset has 3 subgroups with treatment interactions. The first subgroup consists of observations with duration $\leq 6$ and anxiety $\leq 10$. In this subgroup, the mean of the response variable for Treatment 1 is 7, and the mean for the response variable for Treatment 2 is 11. The second subgroup consists of observations with duration $\leq 6$ and anxiety $> 10$. In this subgroup, the mean value of the response variable for Treatment 1 and 2 is 9. The third subgroup consists of observations with duration $>6$. In this subgroup, the mean value of the response variable for Treatment 1 is 12, and the mean of the response variable for Treatment 2 is 7.

Observations were drawn from one of ten clusters, each with a different, cluster-specific (i..e, random) intercept. Data was generated such, that covariates and cluster-specific intercepts were uncorrelated. Also, 43\% of variance in posttreatment depression scores was due to treatment-subgroup interactions, and 8\% of variance was due to cluster-specific variation.

@@

\subsection{GLM}
In a clinical trial, where the outcomes of two or more treatments are compared, an overall GLM may be used to estimate treatment effects. The goal is to estimate a model for predicting the value of a treatment outcome $Y$, which follows, for example, a normal, binomial or Poisson distribution \footnote{ an overview of notation used is provided in the Appendix}:

\begin{equation}
\label{eq:expected value}
	E[y_i | x_i] = \mu_i 
\end{equation}

\begin{equation}
\label{eq:fixedeffects}
	g(\mu_{i}) = x_{i}^{\top}\beta 
\end{equation}

Where $y_i$ is the value of response variable for observation $i$, and $g$ is the link function, characterizing the relationship between the linear predictor $x_{i}^{\top}\beta$ and the mean of the response distribution function. In case of a continuous response variable, $g$ is often as taken as the identity function, and $\mu_{i}$ as the mean of a Gaussian distribution with variance $\sigma_{\epsilon}$. Further, $x_{i}^{\top}$ is a vector of fixed-effects predictor variable values for observation $i$, of which the first element takes a value of 1 for the intercept, and the second element takes the value of a dummy indicator for treatment type. $\beta$ is a vector of fixed-effects regression coefficients, the first element representing the intercept, which is the mean value of the linear predictor in the first treatment group, and the second element representing the slope, which is the mean difference in the linear predictor between the first and second treatment groups.

To keep notation and examples simple, we assume $x_{i}^{\top}$ and $\beta$ to have length 2. That is, the effects of only two treatment conditions are estimated and no additional covariates are included in the GLM. However, additional treatment conditions and covariates can easily be included.

As an example, the estimated GLM of the artificial motivating dataset is graphically represented in Figure \ref{fig:fixedeffects}. The boxplots in Figure \ref{fig:fixedeffects} show the distribution of the posttreatment depression scores in both treatment groups. There seems to be little overall difference in effects of both treatments, as the slope of the regression line is nearly zero. We shall see that this does not necessarily mean that posttreatment depression score and treatment type are unrelated, as the effect of treatment may be moderated by variables not yet included in the model.

\vspace{20pt}
\begin{figure}[!htbp]
    \includegraphics[width=6cm]{glm_example.pdf}
	\caption{Example of a linear fixed-effects model for treatment outcomes, based on the artificial motivating dataset (N=150). The dot for Treatment 1 represents the first, and the slope of the regression line represents the second element of $\beta$.}
	\label{fig:fixedeffects}
\end{figure}

@@

The rationale behind MOB is that a global model for all observations, like the GLM in Equation \ref{eq:expected_value} and \ref{eq:fixedeffects}, may not describe the data well, and when additional covariates are available it may be possible to partition the dataset with respect to these covariates, and find a better model in each cell of the partition \cite{ZeilyHoth08}. This is reminiscent of the classification and regression tree (CART) algorithm of \citeA{BreiyFrie84}, which splits the dataset into subsets, for which the distributions of the outcome variable are most different. However, CART trees detect differences in constant fits across terminal nodes, whereas MOB trees detect differences in parametric models across terminal nodes.

To find partitions and better-fitting local GLMs, the MOB algorithm tests for parameter instability. When the partitioning is based on a GLM, instabilities are differences in $\hat{\beta}$ across partitions of the dataset, which are defined by one or more auxiliary covariates not included in the linear predictor. To find partitions, the MOB algorithm cycles iteratively through the following steps \cite{ZeilyHoth08}: (1) fit the parametric model to the dataset, (2) test for parameter instability over a set of partitioning variables, (3) if there is some overall parameter instability, split the dataset with respect to the variable associated with the highest instability, (4) repeat the procedure in each of the resulting subgroups.

More specifically, in step (2), to test for parameter instability, the so-called \textit{scores} are computed, using the score function. By definition, the empirical scores of all observations in a dataset sum to zero, and when the model is correctly specified, the expected value of the score for each observation is also zero. Under the null hypothesis of parameter stability, the scores do not systematically deviate from the expected value of zero, when the observations are ordered by the values of a potential partitioning variable $U_k$ \cite<c.f., >{MerkyZeil13}. To statistically test whether the scores systematically deviate from zero with respect to variable $U_k$, the class of generalized M-fluctuation tests is used \cite{Zeil05,ZeilyHorn07}.

If the null hypothesis of parameter stability in step (2) can be rejected, that is, if at least one of the partitioning variables $U_{k}$ has a p-value for the M-fluctuation test below the pre-specified significance level $\alpha$, the dataset is partitioned into two subsets in step (3). In step (3), a binary partition is created using $U_{k*}$, the variable with the minimal p-value in step (2). The split point for $U_{k*}$ is selected, by taking the value that minimizes the sum of the values of the objective function in both partitions \cite{ZeilyHoth08}. In step (4), steps (1) through (3) are repeated in each partition, until the null hypothesis of parameter stability can no longer be rejected.

Due to the binary recursive nature of MOB, the resulting partition can be represented as a binary tree. If the partitioning is based on the GLM, the result is a GLM tree, which has a local fixed-effects regression model in every $j$th ($j = 1,...,J$) terminal node of the tree. As a result, in the GLM-tree model, the value for $\beta$ depends on terminal node $j$ in which observation $i$ `falls':

\begin{equation}
\label{eq:fixedeffects_MOB}
	g(\mu_{i}) = x_{i}^{\top}\beta_{j}
\end{equation}

Alternatively, if the recursive subgroup structure (i.e., the partition) were known, the tree could be estimated as a single GLM. The model could then be written: $g(\mu_{i}) = x_{i}^{* \top}\beta^{*}$, where $x_{i}^{*\top}$ are the values of the $2J$ interactions between the subgroups from the tree, and the elements of $x_{i}$. $\beta^{* \top}$ would also have length $2J$, and contain the subgroup-specific fixed-effects coefficients.

Figure \ref{fig:example_mobtree} provides an example of the GLM-tree model in Equation \ref{eq:fixedeffects_MOB}, based on the artificial motivating dataset. By using the three additional covariates (anxiety, duration and age), MOB partitioned the observations into four subgroups, each with a different estimate for $\beta_j$. Age was correctly not detected as a partitioning variable, and the left- and rightmost subgroups are in accordance with the treatment-subgroup interactions as described above. However, the two subgroups in the middle result from a spurious split.

\begin{figure}[!h]
    \includegraphics[width=15cm]{glmtree_example.pdf}
    \caption{Example of a tree representation of model-based recursive partition, based on the artificial motivating dataset. Three additional covariates (anxiety questionnaire score, duration of depressive symptoms at baseline in months and age) were used as potential splitting variables.}
    \label{fig:example_mobtree}
\end{figure}

\subsection{GLMM}
When a dataset contains observations from multiple clusters (e.g., trials, research centers, or individuals in longitudinal datasets), the GLM in Equation \ref{eq:fixedeffects} may be extended to include cluster-specific, or random effects, and the model becomes a GLMM:

\begin{equation}
\label{eq:mixedeffects}
	g(\mu_{i}) = x_{i}^{\top}\beta + z_{i}^{\top}b_{m}
\end{equation}

Where $z_{i}$ is a vector of random-effects predictor variables values for observation $i$, and $b_{m}$ is the vector of random-effects regression coefficients in cluster $m$ ($m=1,...,M$), of which observation $i$ is part. Within the GLMM, it is assumed that $b$ is normally distributed, with mean zero and variance $\sigma^{2}_{b}$. The parameters of the GLMM can be estimated with, for example, maximum likelihood (ML) and restricted ML (REML), as described in \citeA{BrykyRaud92}, for example.

For simplicity, we assume $z_{i}$ and $b_{m}$ to have length 1 in the current paper; that is, only cluster-specific intercepts are included in the models. However, random-effects covariates and coefficients can easily be included. Note that, alternatively, if the random-effects coefficients were known, values of $z_{i}^{\top}b_{m}$ could be included as an offset (i.e., a variable with a fixed coefficient of 1) in the linear predictor of a GLM.

\subsection{GLMM tree}

As noted earlier, ordinary GLM(M)s are not well suited for the detection of treatment-subgroup interactions, whereas the MOB algorithm is, but does not allow for estimation of random effects. Therefore, we propose the GLMM tree, which combines the GLMM from Equation \ref{eq:mixedeffects} with the tree from Equation \ref{eq:fixedeffects_MOB}:

\begin{equation}
\label{eq:glimmertree}
	g(\mu_{i}) = x_{i}^{\top}\beta_{j} + z_{i}^{\top}b_{m}
\end{equation}

To estimate the parameters of this model, we take an approach similar to that of \citeA{HajjyBell11} and \citeA{SelaySimo12}. \citeA{HajjyBell11} and \citeA{SelaySimo12} developed a method for estimation of mixed-effects regression trees (MERTs), which are somewhat similar to GLMM trees. In the MERT approach, the fixed-effects part of a GLMM is replaced by a CART regression tree, and the random-effects part is estimated as usual. To estimate a MERT, an iterative approach is taken, alternating between (1) assuming random effects known, allowing for estimation of the regression tree, and (2) assuming the regression tree known, allowing for estimation of the random effects.

For estimating GLMM trees, we take the MERT approach a step further, by using a GLM tree instead of a regression tree with constant fits. This allows not only for detection of differences in main effects,but also for detection of differences in regression effects (e.g., of treatment type) across terminal nodes. In addition, GLMM trees can be estimated for continuous, as well as binary and count variables. The GLMM-tree algorithm takes the following steps to estimate the model in Equation \ref{eq:glimmertree}:

\vspace{5pt}
\noindent Step 0: Initialize by setting $r$ and all values $\hat{b}_{m(r)}$ to 0.

\vspace{5pt}
\noindent \hangindent=4em Step 1: Set $r = r+1$. Estimate GLM tree ($x_{i}^{\top}\hat{\beta}_{j(r)}$) using responses $y_i - z_{i}^{\top}\hat{b}_{m(r-1)}$.

\vspace{5pt}
\noindent Step 2: Estimate random effects $z_{i}^{\top}\hat{b}_{m(r)}$ using responses $y_i - x_{i}^{\top}\hat{\beta}_{j(r)}$.

\vspace{5pt}
\noindent Step 3: Repeat steps 1 and 2 until convergence.

\vspace{5pt}
The algorithm initializes by setting all values $b_m$ to $0$, since the random-effects (and also the fixed-effects) parts are initially unknown. In every iteration, the GLM tree (i.e., the partition and corresponding fixed-effects coefficients $\beta_j$) and random-effects coefficients $b_m$ are re-estimated. The GLM tree is estimated, given the estimated $b_m$ values from the last iteration, and the $b_m$ values are estimated, given the estimated GLM tree from the current iteration. Iterations are continued until convergence, which is monitored by computing the log-likelihood criterion of the mixed-effects model in Equation \ref{eq:mixedeffects}.
